---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
metadata <- rmarkdown::yaml_front_matter("paper.Rmd")
```

# `r metadata$title`


This repo contains the materials to fully reproduce the paper titled "`r metadata$title`" by `r knitr::combine_words(purrr::map_chr(metadata$author, ~.x$name))`.



## Citation 

Please cite the paper as:


Li et al. (2023) `r metadata$title`. arXiv preprint. URL: http://arxiv.org/abs/2308.05964


## Reproducibility

The main paper is written using R Markdown and contained in "paper.Rmd". To render this, you need to install all relevant packages. Two custom packages are:

```{r, eval = FALSE}
# OOP supports needed by `visage`
remotes::install_github("TengMCing/bandicoot")
# 
# Visual inference models and p-value calculation
remotes::install_url("https://github.com/TengMCing/visage/raw/master/built/visage_0.1.0.tar.gz")
```
